---
title: "07 Case study: comparing Twitter archives"
author: "H. David Shea"
date: "`r format(Sys.time(), '%d %b %Y')`"
output:
  html_document:
    fig.align: center
    fig_caption: yes
    fig_height: 5
    fig_width: 9
    highlight: tango
    theme: united
    #toc: yes
  pdf_document:
    toc: yes
    latex_engine: xelatex
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(knitr)
library(tidytext)
library(scales)

knitr::opts_chunk$set(
    comment = "#>", 
    collapse = TRUE, 
    message = FALSE,
    fig.width = 8,
    fig.asp = ((1 + sqrt(5)) / 2) - 1, # the golden ratio - technically, the b proportion of a+b when a is 1
    out.width = "70%",
    fig.align = "center"
)
```

## Getting the data and distribution of tweets

```{r c7.1a, include=TRUE, warning=FALSE, fig.cap="All tweets from the authors' (Julia Silge and David Robinson) accounts"}
tweets_julia <- read_csv("data/tweets_julia.csv")
tweets_dave <- read_csv("data/tweets_dave.csv")
tweets <- bind_rows(tweets_julia %>%
                        mutate(person = "Julia"),
                    tweets_dave %>%
                        mutate(person = "David")) %>%
    mutate(timestamp = ymd_hms(timestamp))

ggplot(tweets, aes(x = timestamp, fill = person)) +
    geom_histogram(position = "identity",
                   bins = 20,
                   show.legend = FALSE) +
    facet_wrap( ~ person, ncol = 1) +
    theme_light()
```

## Word frequencies

Cleaning up tweet into text only words

```{r c7.2a, include=TRUE, warning=FALSE}
remove_reg <- "&amp;|&lt;|&gt;"
tidy_tweets <- tweets %>%
    filter(!str_detect(text, "^RT")) %>% # remove re-tweets
    mutate(text = str_remove_all(text, remove_reg)) %>% # remove hypertext characters
    unnest_tokens(word, text, token = "tweets") %>%
    filter(
        !word %in% stop_words$word,
        !word %in% str_remove_all(stop_words$word, "'"),
        str_detect(word, "[a-z]")
    )
```

Word frequencies

```{r c7.2b, include=TRUE, fig.cap="Comparing the frequency of words used by Julia and David", fig.width=8, fig.height=8, fig.asp=1}
frequency <- tidy_tweets %>%
    group_by(person) %>%
    count(word, sort = TRUE) %>%
    left_join(tidy_tweets %>%
                  group_by(person) %>%
                  summarise(total = n()),
              by = "person") %>%
    mutate(freq = n / total)

frequency %>%
    slice_head(n = 5) %>%
    kable()

frequency <- frequency %>%
    select(person, word, freq) %>%
    pivot_wider(names_from = person, values_from = freq) %>%
    arrange(Julia, David)

frequency %>%
    slice_head(n = 10) %>%
    kable()

ggplot(frequency, aes(Julia, David)) +
    geom_jitter(
        alpha = 0.1,
        size = 2.5,
        width = 0.25,
        height = 0.25
    ) +
    geom_text(aes(label = word), check_overlap = TRUE, vjust = 1.5) +
    scale_x_log10(labels = percent_format()) +
    scale_y_log10(labels = percent_format()) +
    geom_abline(color = "red") +
    theme_light()
```

## Comparing word usage

Calculate the log odds ratio between David and Julia.

$$\text{log odds ratio} = \ln{\left(\frac{\left[\frac{n+1}{\text{total}+1}\right]_\text{David}}{\left[\frac{n+1}{\text{total}+1}\right]_\text{Julia}}\right)}$$
```{r c7.3a, include=TRUE}
tidy_tweets <- tidy_tweets %>% # Just look at 2016
    filter(timestamp >= as.Date("2016-01-01"),
           timestamp < as.Date("2017-01-01"))

word_ratios <- tidy_tweets %>%
    filter(!str_detect(word, "^@")) %>%
    count(word, person) %>%
    group_by(word) %>%
    filter(sum(n) >= 10) %>%
    ungroup() %>%
    pivot_wider(names_from = person,
                values_from = n,
                values_fill = 0) %>%
    mutate_if(is.numeric, list( ~ (. + 1) / (sum(.) + 1))) %>%
    mutate(logratio = log(David / Julia)) %>%
    arrange(desc(logratio))

word_ratios %>%
    arrange(abs(logratio)) %>% 
    slice_head(n = 10) %>% 
    kable(caption = "Words about equally likely to come from David or Julia's account during 2016")
```

```{r c7.3b, include=TRUE, fig.cap="Comparing the odds ratios of words from the authors' accounts"}
word_ratios %>%
    group_by(logratio < 0) %>%
    slice_max(abs(logratio), n = 15) %>%
    ungroup() %>%
    mutate(word = reorder(word, logratio)) %>%
    ggplot(aes(word, logratio, fill = logratio < 0)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    ylab("log odds ratio (David/Julia)") +
    scale_fill_discrete(name = "", labels = c("David", "Julia")) +
    theme_light()
```

## Changes in word use

Which words’ frequencies have changed the fastest in the authors' Twitter feeds?

```{r c7.4a, include=TRUE}
words_by_time <- tidy_tweets %>%
    filter(!str_detect(word, "^@")) %>% # remove user names
    mutate(time_floor = floor_date(timestamp, unit = "1 month")) %>% # measure monthly
    count(time_floor, person, word) %>%
    group_by(person, time_floor) %>%
    mutate(time_total = sum(n)) %>%
    group_by(person, word) %>%
    mutate(word_total = sum(n)) %>%
    ungroup() %>%
    rename(count = n) %>%
    filter(word_total > 30)

words_by_time %>%
    slice_head(n = 10) %>%
    kable(caption = "Data showing a person using a word in a given month")
```

"The `count` column tells us how many times that person used that word in that time bin, the `time_total` column tells us how many words that person used during that time bin, and the `word_total` column tells us how many times that person used that word over the whole year."

```{r c7.4b, include=TRUE, warning=FALSE}
nested_data <- words_by_time %>%
    nest(-word,-person)

nested_data

nested_models <- nested_data %>%
    mutate(models = map(data, ~ glm(cbind(count, time_total) ~ time_floor, ., family = "binomial")
    ))

nested_models

slopes <- nested_models %>%
    mutate(models = map(models, tidy)) %>%
    unnest(cols = c(models)) %>%
    filter(term == "time_floor") %>%
    mutate(adjusted.p.value = p.adjust(p.value))

top_slopes <- slopes %>% 
  filter(adjusted.p.value < 0.05)

top_slopes %>% 
    select(-data) %>% 
    kable(caption = "Words which have changed in frequency at a moderately significant level in the authors' tweets")
```

```{r c7.4c, include=TRUE, fig.cap="Trending words in David’s tweets"}
words_by_time %>%
    inner_join(top_slopes, by = c("word", "person")) %>%
    filter(person == "David") %>%
    ggplot(aes(time_floor, count / time_total, color = word)) +
    geom_line(size = 1.3) +
    labs(x = NULL, y = "Word frequency") +
    theme_light()
```

```{r c7.4d, include=TRUE, fig.cap="Trending words in Julia’s tweets"}
words_by_time %>%
    inner_join(top_slopes, by = c("word", "person")) %>%
    filter(person == "Julia") %>%
    ggplot(aes(time_floor, count / time_total, color = word)) +
    geom_line(size = 1.3) +
    labs(x = NULL, y = "Word frequency") +
    theme_light()
```

## Favorites and retweets

```{r c7.5a, include=TRUE}
tweets_julia <- read_csv("data/juliasilge_tweets.csv")
tweets_dave <- read_csv("data/drob_tweets.csv")
tweets <- bind_rows(tweets_julia %>%
                        mutate(person = "Julia"),
                    tweets_dave %>%
                        mutate(person = "David")) %>%
    mutate(created_at = ymd_hms(created_at))

tidy_tweets <- tweets %>%
    filter(!str_detect(text, "^(RT|@)")) %>% # keep re-tweets and favorites
    mutate(text = str_remove_all(text, remove_reg)) %>% # remove hypertext characters
    unnest_tokens(word, text, token = "tweets", strip_url = TRUE) %>%
    filter(!word %in% stop_words$word,
           !word %in% str_remove_all(stop_words$word, "'"))

tidy_tweets %>% 
    slice_head(n = 10) %>% 
    kable()

totals <- tidy_tweets %>% 
  group_by(person, id) %>% 
  summarise(rts = first(retweets)) %>% 
  group_by(person) %>% 
  summarise(total_rts = sum(rts))

totals %>%
    kable()

word_by_rts <- tidy_tweets %>%
    group_by(id, word, person) %>%
    summarise(rts = first(retweets)) %>%
    group_by(person, word) %>%
    summarise(retweets = median(rts), uses = n()) %>%
    left_join(totals) %>%
    filter(retweets != 0) %>%
    ungroup()

word_by_rts %>%
    filter(uses >= 5) %>%
    arrange(desc(retweets)) %>% 
    slice_max(retweets, n = 10) %>% 
    kable()
```

```{r c7.5b, include=TRUE, fig.cap="Words with highest median retweets"}
word_by_rts %>%
    filter(uses >= 5) %>%
    group_by(person) %>%
    slice_max(retweets, n = 10) %>%
    arrange(retweets) %>%
    ungroup() %>%
    mutate(word = factor(word, unique(word))) %>%
    ungroup() %>%
    ggplot(aes(word, retweets, fill = person)) +
    geom_col(show.legend = FALSE) +
    facet_wrap( ~ person, scales = "free", ncol = 2) +
    coord_flip() +
    labs(x = NULL,
         y = "Median # of retweets for tweets containing each word") +
    theme_light()
```

```{r c7.5c, include=TRUE, fig.cap="Words with highest median favorites"}
totals <- tidy_tweets %>%
    group_by(person, id) %>%
    summarise(favs = first(favorites)) %>%
    group_by(person) %>%
    summarise(total_favs = sum(favs))

word_by_favs <- tidy_tweets %>%
    group_by(id, word, person) %>%
    summarise(favs = first(favorites)) %>%
    group_by(person, word) %>%
    summarise(favorites = median(favs), uses = n()) %>%
    left_join(totals) %>%
    filter(favorites != 0) %>%
    ungroup()

word_by_favs %>%
    filter(uses >= 5) %>%
    group_by(person) %>%
    slice_max(favorites, n = 10) %>%
    arrange(favorites) %>%
    ungroup() %>%
    mutate(word = factor(word, unique(word))) %>%
    ungroup() %>%
    ggplot(aes(word, favorites, fill = person)) +
    geom_col(show.legend = FALSE) +
    facet_wrap( ~ person, scales = "free", ncol = 2) +
    coord_flip() +
    labs(x = NULL,
         y = "Median # of favorites for tweets containing each word") +
    theme_light()
```

"In general, the same words that lead to retweets lead to favorites."
